UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2018-12-06 ms/main.tex
edited discussion of convergence

\adpcomment{Added this subsection}
\subsubsection*{Performance of t-SNE and UMAP}
To compare UMAP to t-SNE, we ran Python's sklearn implementation of t-SNE on the top 10 PCs of each dataset, once using the default initialization, and once initializing it with a two-dimensional UMAP embedding and skipping the early exaggeration phase of t-SNE (typically the first 250 iterations). The two approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}). Using the UKBB, it becomes clear that UMAP is far better equipped to handle massive genomic datasets with individuals from a variety of ancestral populations. Looking at the Kullback-Leibler divergence after 20,000 iterations, t-SNE's default initialization did not improve upon initialization with the UMAP embedding (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}). Qualitatively the standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but broke the global continuity (figure~\ref{fig:supp_tsne_umap_compare_ukbb}). Given the many desirable properties of UMAP, such computational performance and intuitive placement of individuals within a global genetic continuum, we conclude that UMAP is a superior alternative to t-SNE for large and diverse genomic datasets.

>>>>

\adpcomment{Added this subsection}
\subsubsection*{Comparing t-SNE and UMAP}
Assessing the relative merits of data visualization methods is notoriously difficult, because different users might give different importance to different features of the data. Methods differ in their choice of a similarity metric between high- and low-dimensional representation, which are hard to assess, 
 but also in their strategy to optimize the metrics, which are amenable to direct comparison. This is especially true with UMAP and t-SNE which, despite different mathematical motivations, have computationally similar similarity matrices. Given that t-SNE generated a smattering of clusters for the minority populations when applied to the UKBB, we hypothesized that this was due to a convergence issue rather than a property of the global optimum of the t-SNE metric.

To compare UMAP to t-SNE directly and assess the role of convergence, we ran UMAP and Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a UMAP run with \sgcomment{x epochs}. The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}). When applied to the UKbiobank, we find (perhaps subjectively) that UMAP provides a more compelling visualization of the data compared to both version of t-SNE. We also find that the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), confirming that the default t-SNE optimization fails to identify the global t-SNE minimum. Qualitatively, the standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but broke the global continuity between clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}). Given the many desirable properties of UMAP, such as computational performance and intuitive placement of individuals within a global genetic continuum, we conclude that UMAP is a superior alternative to t-SNE for large and diverse genomic datasets.
